### 2022-11-08 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

\documentclass[12pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs}
\usepackage[colorlinks, allcolors = blue]{hyperref}
\usepackage{natbib}
\citestyle{egu}

@@

\renewcommand{\subsectionautorefname}{Section}
\renewcommand{\subsubsectionautorefname}{Section}

\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
\setcounter{footnote}{1}

@@

@@

Because the mutation rate is biased (\autoref{mutbias}), the probability that a mutation does not occur in a DNA sequence depends on the composition of the sequence.

The probability that an \texttt{A} or a \texttt{T} mutates ($\lambda_\texttt{AT}$), is described by the mutation bias (\autoref{mutbias}) as:

@@

The consensus sequence of TATA-box is \texttt{TATA [T/A] A [T/A] [G/A]} \citep{tata2}. Square brackets mean that, any of the nucleotides enclosed by them, can exist in that specific position. Another study used X-ray crystallography to identify mutations that are tolerated in TATA-box consensus sequence \cite{tata1}. Thus there are eight consensus sequences, and at seven positions, a non-consensus nucleotide is tolerated (\autoref{tatatab}). This makes a TATA-box feature set that contains 136 sequences. We defined the corresponding non-feature set by generating all 8-mer DNA sequences that do not belong to the TATA-box feature set.

@@

\paragraph{Inr:} The consensus sequence of Inr is \texttt{[T/G/C] [T/G/C] CA [T/G/C] [T/A]} \citep{Inr}, which gives 54 sequences in the Inr feature set. We generated the corresponding non-feature set by taking all DNA 6-mers that are distinct from the 54 Inr sequences.

@@

$$Q_\textit{TATA-only} = \frac{P_\textit{TATA}\times(1 - P_\textit{Inr})}{(P_\textit{TATA} + P_\textit{Inr} - P_\textit{TATA}\times P_\textit{Inr})}$$

@@

\vspace{-1em}

\begin{align*}
Q_\textit{Inr-only} & = \frac{P_\textit{Inr}\times(1 - P_\textit{Inr})}{(P_\textit{TATA} + P_\textit{Inr} - P_\textit{TATA}\times P_\textit{Inr})} \\[1em]
Q_\textit{both} & = 1 - Q_\textit{TATA-only} -Q_\textit{Inr-only}
\end{align*}

@@

\begin{align}
P_\textit{RNA-loss} & = Q_\textit{TATA-only} \times P_\textit{TATA-loss}\times(1-P_\textit{Inr-gain}) \nonumber\\[1ex]
 & \quad + Q_\textit{Inr-only}\times P_\textit{Inr-loss}\times(1-P_\textit{TATA-gain}) \nonumber \\[1ex]
 & \quad + Q_\textit{both}\times P_\textit{TATA-loss} \times P_\textit{Inr-loss} + P_\textit{polyA-loss}
\end{align}

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec}
\usepackage[colorlinks, allcolors = blue]{hyperref}
\usepackage{natbib}
\citestyle{egu}

@@

\renewcommand{\subsectionautorefname}{Section}
\renewcommand{\subsubsectionautorefname}{Section}
\titleformat{\paragraph}[hang]{\normalfont\normalsize\itshape}{\theparagraph}{0em}{}
\titlespacing*{\paragraph}{0pt}{1\baselineskip}{3pt}
\setlength{\parindent}{0pt}

@@

\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{1em}

@@

Because the mutation rate is biased (\autoref{mutbias}), the probability that no mutation occurs in a DNA sequence depends on its composition.

The probability that an \texttt{A} or a \texttt{T} mutates ($\lambda_\texttt{AT}$), is thus described as:

@@

The consensus sequence of TATA-box is \texttt{TATA [T/A] A [T/A] [G/A]} \citep{tata2} such that, any of the nucleotides enclosed within the square brackets, can exist in that specific position in the sequence. Another study used X-ray crystallography to identify mutations that are tolerated in TATA-box consensus sequence \citep{tata1}. Thus there are eight consensus sequences, and at seven positions, a non-consensus nucleotide is tolerated (\autoref{tatatab}). This makes a TATA-box feature set that contains 136 sequences. We defined the corresponding non-feature set by generating all 8-mer DNA sequences that do not belong to the TATA-box feature set.

@@

\paragraph{Inr:} The consensus sequence of Inr is \texttt{[T/G/C] [T/G/C] CA [T/G/C] [T/A]}, which gives 54 sequences in the Inr feature set \citep{Inr}. We generated the corresponding non-feature set with all DNA 6-mers that are distinct from the 54 Inr sequences.

@@

\begin{equation*}
Q_\textit{TATA-only} = \frac{P_\textit{TATA}\times(1 - P_\textit{Inr})}{(P_\textit{TATA} + P_\textit{Inr} - P_\textit{TATA}\times P_\textit{Inr})}
\end{equation*}

\vspace{1ex}

@@

\begin{align*}
Q_\textit{Inr-only} & = \frac{P_\textit{Inr}\times(1 - P_\textit{Inr})}{(P_\textit{TATA} + P_\textit{Inr} - P_\textit{TATA}\times P_\textit{Inr})} \\[1ex]
Q_\textit{both} & = 1 - Q_\textit{TATA-only} -Q_\textit{Inr-only}
\end{align*}

@@

\begin{align}
P_\textit{RNA-loss} & = Q_\textit{TATA-only} \times P_\textit{TATA-loss}\times(1-P_\textit{Inr-gain}) \nonumber\\[1pt]
 & \quad + Q_\textit{Inr-only}\times P_\textit{Inr-loss}\times(1-P_\textit{TATA-gain}) \nonumber \\[1pt]
 & \quad + Q_\textit{both}\times P_\textit{TATA-loss} \times P_\textit{Inr-loss} + P_\textit{polyA-loss}
\end{align}

\subsection{Probabilities of finding, gaining, and losing an ORF}

\subsubsection{Probability of finding an ORF}

An open reading frame (ORF) is a nucleotide sequence with a length that is a multiple of three, that begins with a start codon (\texttt{ATG}), and ends with one of the three stop codons (\texttt{TAG}, \texttt{TGA}, \texttt{TAA}). This necessarily means that there are no stop codons within the sequence. Thus the probability of finding an ORF containing $k$ codons including start and stop codons ($P_\textit{ORF}$) is:

\begin{equation}
P_\textit{ORF} = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
\end{equation}

Here, $P_\textit{ATG}$ and $P_\textit{stop}$ are the probabilities of finding a start codon, and a stop codon by chance, respectively.

\subsubsection{Probability of gaining an ORF}

As we defined in the previous section, an ORF has three requirements (start codon, stop codon, and no premature stop codon in the sequence). Thus an ORF can emerge due to mutations via three mechanisms. In each of these mechanisms one requirement is absent whereas the other two are present.
